---
title: "From Excel to R"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(knitr)
library(dplyr)
#remotes::install_github("rstudio/gradethis")
gradethis::gradethis_setup()
```

# Introduction to R data types, variables and functions

## Introduction

### Why should you learn to code if you work with spreadsheets ?

Many people who use computers for their job or studies work with some from of spreadsheets like Excel or Google Sheets. Spreadsheets are a handy way to collect and organise many types of data. Tools like Excel provide additional clever functionality that can help you analyse the data and gain new insights. However, when the dataset is complex, gets large, or the analysis becomes more customized, manually curating and analyzing the data becomes difficult. This often is the point where it is handed over to a data analyst, who will use programming languages to automate the process and perform the seemingly hard tasks. 

Programming has come a long way and learning how to use it for specific tasks is less daunting that it might seem. In this tutorial, we'll introduce you to one of the programming languages that's popular in data science and analytics called _R_. By following along in this interactive page, we'll introduce some very basic data types in R and show how you can use them to construct something very similar to an Excel spreadsheet. This will prepare you for the next step: learning to analyze data from spreadsheet using simple R code.

### Tutorial setup

As mentioned, this tutorial is interactive. You will be able to play around with actual R code provided in the coding boxes like the one below. You can type and run code to try things for yourself.

```{r demo, exercise=TRUE}
"Hello World!"
```
```{r demo-check}
grade_result(
  pass_if(~identical(.result, "Hello World!"))
)
```

- To run code, click the `Run Code` button. You will see any output appear below the box
- In some cases, there is a `Submit Answer` button that will provide feedback
- If you make mistakes you will see errors appear, but don't worry, there is a `Start Over` button in case you need it.

The tutorial is broken up in several topics, and your progress is automatically saved (as long as you don't wipe the browser history). You can also choose to reset the whole tutorial by clicking the small `start over` link underneath the topic overview. 

Ready? Click the `Next Topic` button to get started ...
___

_This tutorial was made to prepare students for an in-person workshop series on coding with R for people with no prior experience but who perform data analysis using Excel. It is not considered a comprehensive introduction to programming, but rather focusses on aspects relevant to the workshop series. In case of questions, please reach out to PJ Van Camp - pjvancamp@hms.harvard.edu_

## Data Types

There are many different types of data we put into spreadsheets, but in most cases we can break them down into a few simple categories of values. Three of the most common types of data entered into a spreadsheet are _numbers_, _text_ and _boolean_ values.

### Numbers

In most programming languages like R, you can work with numbers the same way as you would in a calculator.
For example:
```{r ex1, exercise=TRUE, exercise.eval=TRUE}
1 + 1
```

_Try any of the other basic arithmetic operations using `-`, `*` or `/` and try using decimal numbers as well like `2.5`_

_--- SIDE NOTE ---_

_In programming things are always very explicit, and whole numbers (integers), decimal numbers, complex numbers etc. are technically all considered different data types, though for this intro we won't worry about that._


### Text

The other common type of values are text (often called _strings_ when programming). 
You can write strings by wrapping your text in quotes like this:
```{r ex2, exercise=TRUE, exercise.eval=TRUE}
"This is a string of text"
```
 _Try changing the text_
 
### Boolean

A Boolean value is a binary value that simply means _true_ or _false_. In R these can be written as `TRUE` / `FALSE` or `T` / `F`
```{r ex3, exercise=TRUE, exercise.eval=TRUE}
TRUE
```
_Note that this is case sensitve and you must use all capital letters_
 
### Vectors
 
We can group multiple values of the _same type_ by wrapping them in a vector `c()`

```{r ex4, exercise=TRUE}
c(1,2,3,4,5)
```
_Replace the vector with one containing 3 strings: cat, dog, horse_ 

_You can check your answer once finished_
```{r ex4-check}
grade_result(
  pass_if(~identical(.result, c("cat", "dog", "horse")))
)
```

### Lists
 
We can group multiple values of the _different types_ by creating a list `list()`

```{r ex5, exercise=TRUE}
list(TRUE, "test", c(1,2,3))
```
_This list has 3 entries: a boolean, a string and a vector (which in turn contains 3 numbers)_

Lists allow you to give a unique name to each of its entries like so:

```{r ex6, exercise=TRUE}
list(item1 = TRUE, item2 = "test", item3 = c(1,2,3))
```

### Note on _lines of code_

You can see that the code boxes have line numbers. This is because when the code is run, it's done so line by line (each line containing a new instruction). However, there are ways of splitting a long line of code across multiple lines for better readability without _braking_ the code. 

Examples of valid ways to split a line are _after_ an opening bracket `(` or a comma `,` or _before_ a closing bracket `)`. When R executes the code, it will ignore these line breaks and interpret them as if they were on as single line. This means we can rewrite ...
```r
list(item1 = TRUE, item2 = "test", item3 = c(1,2,3))
```
... like so ...

```r
list(
  item1 = TRUE, 
  item2 = "test", 
  item3 = c(1,2,3)
)
```
... and it will still be valid and interpreted as a single line.

### Try it yourself

Create the following named list:

- The first entry contains a vector of the numbers 1 to 3 and is named _first_
- The second entry contains a vector with the first 3 letters of the alphabet (lowercase) and is named _second_

```{r ex7, exercise=TRUE}
list()
```


```{r ex7-check}
grade_result(
  pass_if(~identical(.result, list(first = c(1,2,3), 
                                   second = c("a", "b", "c"))
))
)
```

## Data Frames

### Building on what we know ...

In the previous topic you learned about basic R data types. Let's now use then to recreate a spreadsheet. Start by taking a look at this spreadsheet table below

```{r echo=FALSE}
data.frame(name = c("Mohamet", "Lin Yi", "Nina"), age = c(21,19,24),
           student = c(TRUE, TRUE, FALSE)) %>% kable()
```

This simple spreadsheet records student info and is a table with the following properties:

- Each row is an entry in the table (i.e. info about a particular student)
- Each column records a different observation
- Every column has a header that describes what is being recorded

Note that the values in every column are of the same data type. This means we can represent them as a vector. For example, the first column would be `c("Mohamet", "Lin Yi", "Nina")`. If we do this for all three columns and organise them as a list in R, we can recreate the whole table like this:

```{r}
list(
  name = c("Mohamet", "Lin Yi", "Nina"), 
  age = c(42,33,19),
  student = c(TRUE, TRUE, FALSE)
) 
```

We just converted our spreadsheet to a combination of data types in R! Note that the way it written in R looks like the table was rotated with the columns now horizontal an the rows vertical.

### Introducing the R Data Frame

Although the list above is a valid representation of a table, R has another, special, built-in datatype that optimises working with this type of data called a _data frame_. It is constructed very much like we did with the list but is handled and displayed differently:
```{r}
data.frame(
  name = c("Mohamet", "Lin Yi", "Nina"), 
  age = c(42,33,19),
  student = c(TRUE, TRUE, FALSE)
) 
```

A data frame imposes more restrictions on how the data can be organised as it has to conform to the rules of a well-organised spreadsheet table. These rules are:

- Every column needs a unique name (avoid using spaces)
- Every column can only hold a single data type (i.e. need to be represented as a vector)
- The length of each column has to be the same

_There are more rules that you should follow to create a tidy table, but they will not be discussed in here_

### Try it yourself

Edit the following code so it becomes a _valid_ data frame with 3 rows, and the columns: animal, legs and mammal. 

```{r ex8, exercise=TRUE}
data.frame(
  animal = c("dog", "beetle", "spider"), 
  mammal = c(TRUE),
  c(4, 6, 8)
) 
```
_You will see errors appear as long as it's not correct, but keep trying!_

```{r ex8-check}
grade_result(
  pass_if(~identical(.result, data.frame(
    animal = c("dog", "beetle", "spider"), 
    mammal = c(TRUE, FALSE, FALSE),
    legs = c(4, 6, 8)
    ) 
))
)
```

## Variables

Now we know how to create different types of data in R, let's see how we can store and access them in R.

### Storing data into a variable

We define (assign) a variable by picking a name followed by the assignment `<-` operator (`<` followed by `-`) and then the value (or calculation that will generate one).

Valid variable names:

- Must start with a letter
- can only have alphanumeric characters (A-Z, a-z, 0-9), `_` or `.`

```{r ex9, exercise=TRUE, exercise.eval=TRUE}
x <- FALSE
y2 <- data.frame(x = c(1,2,3), y = c("A", "B", "C"))
other_var <- (6 + 4) / 2

x
```

_You can print the content of a variable by simply wiring its name on a separate line (e.g. `x` on line 5). Try and replace it with any of the other variable names to print its content when you run the code_

### Updating a variable

You can update a variable by using the same name, but assigning a different value
```{r ex10, exercise=TRUE}
x <- 1
x
x <- 2 + 3
x
```

We can also use the values stored in variables to perform calculations
```{r ex11, exercise=TRUE}
x <- 5
y <- x * 2
y
```

We can even update a variable by using it's own value then reassigning it
```{r ex12, exercise=TRUE}
x <- 10 - 4
x <- x / 2
x
```

### Try it yourself
Create a variable `prod` that stores the product of x, y and z and then display the result

```{r ex13, exercise=TRUE, exercise.reveal_solution = FALSE}
x <- 5
y <- 7
z <- 11

```


```{r ex13-solution}
x <- 5
y <- 7
z <- 11
prod <- x * y * z
prod
```

```{r ex13-check}
grade_this_code()
```

## Functions

### Intro

The power of programming lies in the ability to execute a set of instructions automatically and being able to reuse code written by others or yourself. This a accomplished using _functions_. Just as in mathematics, a function contains a set of instructions, with variables or parameters that serve as placeholders. 

Let say we have a vector x  ...
```r
x = c(1, 15, 22, 10)
```
... and we want to calculate the mean value. R has a built-in function called `mean` which takes in a vector and returns the mean:
```r
mean(x)
```
```{r ex14, exercise=TRUE, exercise.eval=TRUE}
values <- c(1, 15, 22, 10)
result = mean(values)
result
```

_Try to replace the mean function with some other basic functions you can execute on vectors of numbers like `min` or `max` or `sum`_

### Arguments

Some functions have multiple parameters or arguments that you have to specify. 

For example, the `round` function has two parameters: `x`, which is the value we like to round, and `digits`, the number of decimal points we want to round to. If we want to round a number to 3 decimal points we specify them inside the function ...
```{r}
value <- 17/13
value
round(x = value, digits = 3)
```
... we set `x` to the value and `digits` to 3.

To know which arguments a function takes, you have several options

#### Option 1: look at the function code

Run the function _without_ the  brackets `()` to print its code and look at the arguments listed at the top
```{r}
round
```

_Note that the second argument here already has a default value assigned (see notes below for more details)_

#### Option 2: load the function documentation

Run the function _without_ the  brackets and put a question mark in front of it. This will open up the documentation (might not work in this online tutorial) 
```{r ex15, exercise=TRUE}
?round
```
_Note that the documentation has much more details on how to use the function and what each argument means_

#### Option 3: Code editor tooltips

Many code editors like [R Studio](https://posit.co/download/rstudio-desktop/) or [VS Code](https://code.visualstudio.com/) will automatically show function arguments whilst writing code or when you hover over a function name in your code. Even here you will see suggestions pop-up when you start typing the brackets of a known function

```{r ex16, exercise=TRUE}
round

```

#### Notes about function arguments

If you don't specify the argument names explicitly, the first value will become the first listed argument, the second the next, etc.
```{r}
round(1.333, 1)
```
_This implicit argument assignment will be translated to round(x = 1.333, digits = 1) when the code is run_

Functions can come with default values for some of its arguments, which you don't have to set if you are OK with the value. In other words, functions have required arguments (i.e. no default and you have to provide a value) and optional arguments (you can choose to override the default if you want). Look at the function documentation to see which ones are required.
```{r}
round(1.333)
```
_By default `digits = 0`_

You can combine implicit (unnamed) and explicit (named) arguments, but the implicit ones always have to come first and all required arguments need to have a value. Let's take the [format](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/format) function as an example
```{r}
format(15658475, big.mark = ",")
```
_In this example, big.mark is the 9th argument, but since all apart from the first argument have default values, we can use named arguments to only override those we need to change_

### Try it yourself

1. You are given a variable _values_ that contains a vector of numbers. 
2. On line 2 use the function `sort` to sort the values in the vector and save the result to a new variable called _values_sorted_
3. Now make sure that the sorting happens in _decreasing_ order (i.e. highest to lowest). You will need to set an argument of the sort function to make this happen (look it up)
4. Print the result of _values_sorted_ on line 3

```{r ex17, exercise=TRUE, exercise.reveal_solution = FALSE}
values <- c(11.2, 16, 0.6, 8.89)

```


```{r ex17-solution}
values <- c(11.2, 16, 0.6, 8.89)
values_sorted <- sort(x = values, decreasing = TRUE)
values_sorted
```

```{r ex17-check}
grade_this_code()
```

### Combining functions

Functions can use other functions as input to create more complex behaviour. There are two ways of doing this: 

#### Nested functions

Just as in mathematics, the inner most calculations of nested functions are performed before the outer ones can use their results.

```{r}
floor(sqrt(2))
```
_The square root function is run first before the value is used in the flooring function_

#### Piping (chaining) functions

Highly nested functions can become difficult to read, so it it also possible to write your code differently by using the result from one function as the input for the next one using the piping operator `|>` (`|` followed by `>`).

```{r}
sqrt(2) |> floor()
```

* The function that needs to be executed first comes first in the chain
* The result from the previous function will occupy the _first_ argument slot of the next function. 
* You can add (optional) named arguments to any function in the chain

More complex example:
```{r}
sqrt(2) |> exp() |> round(digits = 3) 
```
_Take the square root of 2, exponentiate it (raise e, Euler's number to this power) and then round the result to 3 digits_

_This is more readable than the nested version `round(exp(sqrt(2)),digits = 3)`_

___

Note on the piping operator

The built-in piping operator `|>` was only introduced in R version 4.1.0 in 2021. It was already possible long before that to use piping but this required an additional package to be installed. Hence, you will still see a lot of code using an alternative pipe which looks like `%>%`. Both are identical in functionality.

### Try it yourself

- On line 2, write a _nested_ set of functions what will `sum` _vec_ and then `floor` the result (i.e. round down to the nearest integer) 
- On line 3, write the same set of functions but now using pipes


```{r ex18, exercise=TRUE, exercise.reveal_solution = FALSE}
vec <- c(65.001,58,412.4,36,87.6666,4,11,0.69)
```


```{r ex18-solution}
vec <- c(65.001,58,412.4,36,87.6666,4,11,0.69)
floor(sum(vec))
sum(vec) |> floor()
```

```{r ex18-check}
grade_this_code()
```

## Vector operations

Remember that a data frame in essence is a list of column vectors. Before we learn to work with this, it's important to know how R makes it very easy to perform operations across vectors. We'll just explore the basics you need here.

### Simple vector math

When you perform basic arithmetic operations on a vector, they will be carried out for every element in that vector and return a new vector with all results. Note that this will onyl work if the vectors are of the same size. 

```{r ex19, exercise=TRUE, exercise.eval=TRUE}
vec <- c(2,4,6,8,10)
vec / 2
```
_You can try out other basic operations like `+`, `-` or `*`_

### Operations with multiple vecors

When you perform basic arithmetic operations using two vectors they will be carried out by using the first element of each vector for the first calculation, the the second for the next one and so on, returning again a vector with all results.

```{r ex20, exercise=TRUE, exercise.eval=TRUE}
vec <- c(1,2,3,4,5)
vec + vec
```
_You can try out other basic operations like `+`, `-` or `*`_

### Functions across vectors

Finally, you can use functions across a vector. The result depends on what the effect of the function is. For example:

```{r}
vec <- c(1.554,6.23,2.201,0.01366)
sum(vec)
```
_As seen before, this will sum all elements in the vector and return a single value_

```{r}
vec <- c(1.554,6.23,2.201,0.1366)
round(vec, digits = 1)
```
_On the other hand, if a function is made to be run on a single value, it will just be repeated for every value in the vector and return the result vector_

### Try it yourself

When _centering_ a set of numbers, you ensure that the mean value is zero without changing the distribution. We can do this by simply subtracting the mean from each individual value. For example: the list 2, 4, 9 has a mean value of 5. If we subtract this from each value we get -3, -1 , 4 which has a mean of 0.

1. You are given a variable _vec_ that contains a set of numbers
2. Calculate the mean value using the `mean` function
3. Use vector operations to subtract the mean from each entry in the vector

```{r ex21, exercise=TRUE, exercise.reveal_solution = FALSE}
vec <- c(11,6,3,8,9)

```
_Note, you can accomplish step 2 and 3 in a single line, or create an intermediate variable. You should try both and get the same result_

```{r ex21-solution}
vec <- c(11,6,3,8,9)
vec - mean(vec)
```

```{r ex21-check}
vec <- c(11,6,3,8,9)
grade_result(
  pass_if(~identical(.result, vec - mean(vec))
))
```

## Wrap-up

Congratulations, you now know all the basics to be ready to start working with data frames in R! You can revisit this introduction at any time, but don't worry if some things are still unclear, learning to code takes time and frustration is definitely part of the experience, but you'll have more opportunity to practice soon! 