---
title: "From Excel to R"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
#remotes::install_github("rstudio/gradethis")
gradethis::gradethis_setup()
```

# Values, Variables and Functions

## Values

Think of values as what goes in a cell within a specific column in you Excel document. The two most common types are _text_ and _numbers_

### Numbers

In R, you can work with numbers the same way as you would in a calculator.
For example:
```{r ex1, exercise=TRUE, exercise.eval=TRUE}
1 + 1
```

_Try any of the other basic numeric operations using `-`, `*` or `/` and try using decimal numbers as well like `2.5`_


### Text

The other common type of values are text (often called _strings_ in code). 
You can write this by wrapping your text in quotes like this:
```{r ex2, exercise=TRUE, exercise.eval=TRUE}
"This is a string of text"
```
 _Try changing the text_
 
## Variables

We can store values for later use inside a _variable_, by defining the variable using a alphanumeric name (no spaces or special characters apart from `_`)

```{r ex3, exercise=TRUE, exercise.eval=TRUE}
x = 1
y = 2
x + y
```
 _Try changing the values of x and y or the operation_

We can store multiple values in a single variable by using a vector `c()`

```{r ex4, exercise=TRUE, exercise.eval=TRUE}
x = c(1,2,3,4,5)
x
```
 _Try changing the values in the vector to a, b, c, d, e_
 
#------------------
```{r ex-check}
grade_result(
  pass_if(~identical(.result, 2))
)
```
