---
title: "From Spreadsheets to R - without going too far"
subtitle: "Part 3 - Plotting your data"
output: html_document
editor_options: 
  chunk_output_type: console
---

*This is an R-markdown document. If run in RStudio, you will see code blocks appear with the option to run each one using the little "play button" on the top-right of each block. The output (including errors) will be printed below a block. For the best view, click the `Visual` mode at the top left of this file (if set to Source now)*

Read the [session materials](README.md) for more background on the various functions and their use

## Make sure you have the following libraries loaded

```{r setup, include=FALSE}
library(knitr)
library(readxl)
library(dplyr)
library(ggplot2)

opts_chunk$set(echo = TRUE)

#install.packages(c("tidyverse", "knitr"))

```

If you get an error that libraries are not installed yet, un-comment the last line (remove the `#`), and run the code block again (you can re-comment once installed).

# Exploring the data

## Read in an Excel file

Let's load in an Excel file derived from a public dataset on water quality measurements.
The dataset contains 500  samples with various measurements that are used to assess the quality of the water (pH, temperature, turbidity, oxygen and conductivity)
<https://www.kaggle.com/datasets/shreyanshverma27/water-quality-testing>

```{r}
water <- read_xlsx("part3_ggplot/Water_Quality.xlsx")
glimpse(water)
```

# Mapping dataframe columns to plot aesthetics

Let's create a graph that contains the measured pH level of each sample on the x-axis and the amount of dissolved oxygen on the y-axis

```{r}
waterPlot <- ggplot(water, aes(x = pH, y = Oxygen))
waterPlot
```

*Note that when the plot is rendered, no data points are shown because we have not defined the exact type of plot yet*

# Visualising data points - geom\_\<type>()

Let's start by plotting the data as points (i.e. a scatter plot)

```{r}
waterPlot + geom_point()
```

We can use the same base ggplot object to plot the data as a line chart instead

```{r}
waterPlot + geom_line()
```
_Note that for this data plotting the data as a line chart does not make sense as all points are independents and thus are not connected in any way_

### Try it yourself

We'll be working with a dataset that comes pre-installed with R called `iris` which has a record of several plant leaf characteristics for different iris species

```{r}
flowers <- iris
glimpse(flowers)
```

#### Make a scatter plot (points) of the petal length vs the width

```{r}
flowersPlot <- ggplot(flowers, aes(x = Petal.Length, y = Petal.Width)) + 
  geom_point()
flowersPlot
```

#### Turn the same plot into a line chart

```{r}
flowersPlot <- ggplot(flowers, aes(x = Petal.Length, y = Petal.Width)) + 
  geom_line()
flowersPlot
```

*Although it renders, why is this not a good chart type for this data?*

#Static vs data-driven attributes

Depending on the data we have and what we want to visualise, we can either specify plot style attributes for all data points at once or have them be driven by the data.

## Global / static attributes

Change the size of all the points in the plot

```{r}
waterPlot <- ggplot(water, aes(x = pH, y = Oxygen)) +
  geom_point(size = 3)
waterPlot
```

*Defining an attribute outside of the aes() function in the geom function will set it for all the data points*

## Data driven point size

Change the size based on the water temperature

```{r}
waterPlot <- ggplot(water, aes(x = pH, y = Oxygen, size = Temp)) +
  geom_point()
waterPlot
```

*Note that the size of the point is now defined by the value of n_actions and a legend appears*

### Try it yourself

#### Continue with the iris flowers scatter plot you created in the previous section and colour all the points blue

```{r}
ggplot(flowers, aes(x = Petal.Length, y = Petal.Width)) + 
  geom_point(color = "blue")
```

*You can either provide the name of the colour like "blue" or a hex value like "#1273ed"*

#### Now change this so that each species has a unique colour

```{r}
ggplot(flowers, aes(x = Petal.Length, y = Petal.Width, colour = Species)) + 
  geom_point()
```

*Tip ggplot will automatically generate unique colours for different text values*

## Plot layout

### Add axis labels and a title - `labs()`

```{r}
waterPlot <- ggplot(water, aes(x = pH, y = Oxygen, size = Temp)) +
  geom_point() +
  labs(x = "Water pH", 
       y = "Dissolved Oxygen (mg/L)", 
       title = "Effect of water pH on oxygenation"
       )
waterPlot
```

*Add axis labels and a title to the plot*

### Change theme - `theme()`

```{r}
waterPlot <- ggplot(water, aes(x = pH, y = Oxygen, size = Temp)) +
  geom_point() +
  labs(x = "Water pH", 
       y = "Dissolved Oxygen (mg/L)", 
       title = "Effect of water pH on oxygenation"
       ) +
  theme_bw()
waterPlot
```

*Use a built-in black and white theme*

### Try it yourself

#### Continue from the last iris scatter plot: update the axis labels and add a title

-   Title: Iris petal length and width comparison for different species
-   x label: Petal length (cm)
-   y label: Petal width (cm)

```{r}
ggplot(flowers, aes(x = Petal.Length, y = Petal.Width, colour = Species)) + 
  geom_point() + 
  labs(x = "Petal length (cm)", y = "Petal width (cm)",
       title = "Iris petal length and width comparison for different species")
```

#### Now add the built-in minimal theme to the plot `theme_minimal()`

```{r}
ggplot(flowers, aes(x = Petal.Length, y = Petal.Width, colour = Species)) + 
  geom_point() + labs(
    x = "Petal length", 
    y = "Petal width",
    title = "Iris petal length and width comparison for different species") + 
  theme_minimal()
```

## Bar charts

When creating bar charts with ggplot you have to summarise the data first if needed so it contains the position and height of each bar.

### Data summary

```{r}
sumData <- water |>
  group_by(pH) |>
  summarise(Oxygen = mean(Oxygen)) |> 
  ungroup()

glimpse(sumData)
```
_We now have a data frame with the average oxygen level for each PH level_

### Create a bar graph

```{r}
waterPlot <- ggplot(sumData, aes(x = pH, y = Oxygen)) +
  geom_bar(stat = "identity")

waterPlot
```

*Note: `stat = "identity"` should always be set when you are creating bar charts where you summarised the data for each bar already*

### Try it yourself

#### Summarise the iris data so you have the average `mean()` petal length per species

```{r}
sumData <- flowers |> 
  group_by(Species) |> 
  summarise(avgPetalLength = mean(Petal.Length)) |> 
  ungroup()
```

*This is a good refresher of Part 2 on working with data frames. Should you really get stuck scroll down to the very end to see the code*

#### Now create a bar chart that shows the average petal length per species

```{r}
ggplot(sumData, aes(x = Species, y = avgPetalLength)) + 
  geom_bar(stat = "identity")
```

#### If your previous plot had horisontal bars, try to plot them vertically now or vice versa in case they were vertical in the previous plot

```{r}
ggplot(sumData, aes(x = avgPetalLength, y = Species)) + 
  geom_bar(stat = "identity")
```

## Saving a plot as an image

Once you are happy with your result, you can save the plot as an image.

```{r}
waterPlot <- ggplot(water, aes(x = pH, y = Oxygen, color = Conductivity)) +
  geom_point(size = 3) +
  labs(x = "Water pH", 
       y = "Dissolved Oxygen (mg/L)", 
       title = "Effect of water pH on oxygenation"
       ) +
  theme_bw()
waterPlot

ggsave(filename = "oxygenation.png", plot = waterPlot, width = 1800, height = 900, units = "px")
```

## Bringing it all together

Let's look at a wine tasting experiment were experts rated red wines on a ten point scale. In addition, specific wine properties were measured.

source: <https://www.kaggle.com/datasets/shilongzhuang/red-and-white-wine-quality>

#### Load the `winequality-red.csv` dataset using the built-in `read.csv()` function into a new variable called `wine` and explore the dataset.

```{r}
wine <- read.csv("part3_ggplot/winequality-red.csv")
glimpse(wine)
```

*Note that this data is not stored as an XLSX file but a CSV file*

#### Create a plot with the following properties:

-   scatter plot
-   fixed acidity on x the axis
-   density on y the axis
-   alcohol level is represented as the point colour
-   all points are of size 1.2
-   title: Correlation between acidity and density in red wines
-   use the built-in light theme

```{r}

ggplot(wine, aes(x = fixed.acidity, y = density, colour = alcohol)) + 
  geom_point(size = 1.2) + theme_light()

```

#### Now summarise the data so that you get the average `mean()` sulphates concentration per quality level and save this to a new varliable `sulQual`

```{r}

sulQual <- wine |> 
  group_by(quality) |> 
  summarise(sulphates = mean(sulphates)) |> 
  ungroup()

```

*Again if you get stuck scroll down to the very end to see the code*

#### Create a bar chart of the results and style the plot as you see fit

```{r}
sulQualPlot <- ggplot(sulQual, aes(x = quality, y = sulphates)) + 
  geom_bar(stat = "identity")
sulQualPlot
```

#### Save the bar chart as a PNG image `wine_sulfate_quality.png` with dimensions 600px x 800px (width x height)

```{r}
ggsave(filename = "wine_sulfate_quality.png",plot =sulQualPlot, width = 600, height = 800, units = "px")
```

## EXTRA - More ggplot functions

Below are some examples of more useful ggplot functions you can explore

### Specify colours

When setting `colour` in the `aes()` function, ggplot will use a default set of colours. You can change this by providing your own colour scale

```{r}
flowers <- iris

flowersPlot <- ggplot(flowers, aes(x = Petal.Length, y = Petal.Width, colour = Species)) + 
  geom_point() + theme_minimal() +
  # Use a discrete set of colours
  scale_colour_discrete(
    name = "Iris species",
    type = c(setosa = "orange", versicolor = "#72BD13", virginica = "purple")
  )
  
flowersPlot
```

*Colours when there are a discrete set of possibilities*

```{r}
flowers <- iris
flowersPlot <- ggplot(flowers, 
                 aes(x = Petal.Length, y = Petal.Width, colour = Sepal.Length)) + 
  geom_point() + theme_minimal() +
  # Use a colour gradient
  scale_colour_gradientn(
    name = "Sepal Length",
    colours = c("#132B43", "#72BD13", "#ED553B")
  )
  
flowersPlot
```

*Set colours across a range of numbers (The colours provided here are equally spaced between min and max)*

## Add text labels

You can add text labels to a plot with `geom_text`

```{r}
flowersPlot <- ggplot(flowers, aes(x = Petal.Length, y = Petal.Width, 
                             label = Species)) + 
  theme_minimal() +
  geom_text(check_overlap = TRUE)
  
flowersPlot
```

*You can set `check_overlap = TRUE` to clean up overlapping text. You can also change to `geom_label()` if you want the labels to have a background colour*

## Add trend line

You a trend line to a set of points with `geom_smooth()`

```{r}
flowersPlot <- ggplot(flowers, aes(x = Petal.Length, y = Petal.Width)) + 
  geom_point() + theme_minimal() +
  geom_smooth(method = "lm", se = FALSE, colour = "#72BD13")
  
flowersPlot
```

*The `method` here is set to a linear model "lm", and standard deviation `se` is hidden*

### Try it yourself

#### Improve the colour scheme of the waterPlot so low conductivity is blue and high is red

```{r}
waterPlot <- ggplot(water, aes(x = pH, y = Oxygen, color = Conductivity)) +
  geom_point(size = 2) +
  labs(x = "Water pH", 
       y = "Dissolved Oxygen (mg/L)", 
       title = "Effect of water pH on oxygenation"
       ) +
  theme_bw() +
  scale_colour_gradientn(
    colours = c("#0044FF", "#ED553B")
  )

waterPlot

```

#### Now add a trend line to the plot

```{r}
waterPlot <- ggplot(water, aes(x = pH, y = Oxygen, color = Conductivity)) +
  geom_point(size = 2) +
  labs(x = "Water pH", 
       y = "Dissolved Oxygen (mg/L)", 
       title = "Effect of water pH on oxygenation"
       ) +
  theme_bw() +
  scale_colour_gradientn(
    colours = c("#0044FF", "#ED553B")
  ) + 
  geom_smooth(method = "lm", se = FALSE, colour = "#72BD13")

waterPlot

```

## SPOILER: Data summary code for exercises

Summarise the iris data so you have the average `mean()` petal length per species

```{r}
sumData <- flowers |> 
  group_by(Species) |> 
  summarise(avgPetalLength = mean(Petal.Length)) |> 
  ungroup()
```

Summarise the wine data so that you get the average `mean()` sulphates concentration per quality level and save this to a new variable `sulQual`

```{r}

sulQual <- wine |> 
  group_by(quality) |> 
  summarise(sulphates = mean(sulphates)) |> 
  ungroup()

```
