---
title: "From Spreadsheets to R - without going too far"
subtitle: "Part 2 - Data Frame Manipulations"
output: html_document
editor_options: 
  chunk_output_type: inline
---

*This is an R-markdown document. If run in RStudio, you will see code blocks appear with the option to run each one using the little "green play button" on the top-right of each block. The output (including errors) will be printed below the block. For the best view, click the `Visual` mode at the top left of this file (if set to Source now)*

Read the [session materials](README.md) for more background on the various functions and their use

## Load the following libraries

```{r setup, include=FALSE}
library(knitr)
library(readxl)
library(dplyr)
library(rstudioapi)

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
opts_chunk$set(echo = TRUE)

# install.packages(c("tidyverse", "knitr"))
```

If you get an error that libraries are not installed yet, un-comment the last line (remove the `#`), and run the code block again (you can re-comment once installed).

# Exploring the data

## Read in an Excel file

Let's load in an Excel file (antiquities_act.xlsx) which is stored in the same folder as this script. We'll use the `read_excel` function and save the data frame to a variable named `data`

```{r}
data <- read_excel("antiquities_act.xlsx")
data
```

*You will see a preview of the table when you run this block, but you can also look at it it by clicking the variable name 'data' in the Environment tab on the right of your screen*

## Handy functions to learn more about the data

The built-in function `colnames()` can list all the column names of a data frame, and the `glimpse()` function from the `dplyr` package will give you a sample of the data in each column

```{r}
colnames(data)
glimpse(data)
```

# Selecting columns - select()

## By name

```{r}
myData <- data |> select(c("current_name", "states", "year"))
glimpse(myData)
```

*Select the columns current_name, states and year*

*Remember that in `dplyr` selecting columns using `select()` function does not require you to use a vector of quoted strings*

```{r}
myData <- data |> select(current_name, states, year)
glimpse(myData)
```

## By index

```{r}
myData <- data |> select(1,2,7)
glimpse(myData)
```

*Select the 1st, 2nd and 7th column*

## Inverted

```{r}
myData <- data |> select(-c(1,2,7))
glimpse(myData)
```

*Select all but the 1st, 2nd and 7th column. Note that in this case a vector is required to invert everything by putting a minus sign in front of it*

### Try it yourself

*Use the same `data` variable used in the examples above for all following exercises. You can save your results to whichever variable you like (e.g. myData)*

#### Select the current agency and the the number of acres affected from the dataset

*Start by looking for the correct column names*

```{r}
myData <- data |> select(current_agency, acres_affected)
```

#### Select the 1st, 2nd and 5th columns

```{r}
myData <- data |> select(1,2,5)
```

#### Select everything but the 2nd and 3rd column

```{r}
myData <- data |> select(-c(2,3))
```

#### Extra challenge: you can select a range of columns using the <first>:<last> notation.

For example 1:5, is the same as 1,2,3,4,5. Using this, select the columns 1-4 and 7-9

```{r}
myData <- data |> select(1:4, 7:9)
```

# Filtering rows - filter()

## Boolean statement

```{r}
myData <- data |> filter(acres_affected < 100)

# Display relevant columns
myData |> select(current_name, acres_affected)
```

*Filter all rows with less than 100 affected acres*

## Multiple boolean statements with *and (&)*

```{r}
myData <- data |> filter(acres_affected < 100 & states == "Florida")

# Display relevant columns
myData |> select(acres_affected, states)
```

*Filter all rows with less than 100 acres affected in Florida*

## Multiple boolean statements with *or (|)*

```{r}
myData <- data |> filter(acres_affected < 100 | acres_affected > 1000000)

# Display relevant columns
myData |> select(current_name, acres_affected)
```

*Filter all rows with less than 100 or more than 1,000,000 affected acres*

### Try it yourself

#### Filter out (i.e keep) all rows from the `data` dataframe from the year 2016 onwards

```{r}
myData <- data |> filter(year >= 2016)
```

*The result should have 13 rows*

#### Filter out all data from *Pinnacles National Park* before 1950

```{r}
myData <- data |> filter(current_name == "Pinnacles National Park" & year < 1950)

```

*The result should have 6 rows*

#### Get all rows with more than 5000 affected acres OR after the year 2000

```{r}
myData <- data |> filter(acres_affected > 5000 | year > 2000) 

```

*The result should have 122 rows*

# Changing columns - mutate()

## New column

```{r}
myData = data |> mutate(years_protected = 2025 - year)

# Display relevant columns
myData |> select(year, years_protected)
```

*Calculate the number of years the land has been protected in 2025*

## Update existing column

```{r}
myData = data |> mutate(acres_affected = ceiling(acres_affected))

# Display relevant columns
myData |> select(current_name, acres_affected)
```

*Round up the acres to the nearest acre*

### Try it yourself

#### Create a new column `large` that stores the boolean value (TRUE / FALSE) to indicate if the number of acres affected is larger than 1000

```{r}
myData <- data |> mutate(large = acres_affected > 1000)
```

#### Use the `toupper()` function to change all state names to be written in capital letters (update the *states* column)

```{r}
myData <- data |> mutate(states = toupper(states))

```

# Summarise data - summarise()

## Summarise everything to a single row

```{r}
data |> 
  summarise(mean_acres = mean(acres_affected), sd_acres = sd(acres_affected))
```

*Summarise the data to show the mean and standard deviation of the affected acres across the whole dataset*

## By group

Use the group_by() function, but don't forget to ungroup() once finished

```{r}
data |> group_by(states) |> 
  summarise(mean_acres = mean(acres_affected), sd_acres = sd(acres_affected)) |>
  ungroup()
```

*Get the mean and standard deviation of the affected acres by state. Note that NA means there is a missing value. In this case that is because there only is one row for those states and you need at least two values to calculate a standard deviation*

### Try it yourself

#### Get the minimum `min()` and maximum `max()` amount of acres affected across the whole table

```{r}
myData <- data |> 
  summarise(min_acres = min(acres_affected), max_acres = max(acres_affected)) |>
  ungroup()
```

#### Do the same thing but now for each year seperately

```{r}
myData <- data |> 
  group_by(year) |>
  summarise(min = min(acres_affected), max = max(acres_affected)) |>
  ungroup()

```

# Putting it all together

In this part you will combine all new things you've learned (selecting, filtering, mutating and summarising) into a more comprehensive analysis

#### Use the `read_excel()` function to load the *antiquities_act.xlsx* file into a new variable called `newData`

```{r}
newData <- read_excel("antiquities_act.xlsx")
```

#### Select the current name, state, action, year and affected acres columns from the data and put them into a new variable called `analysis`

```{r}
analysis <- newData |> select(current_name, states, action, year, acres_affected)
```

#### We'll only focus on the data where and area was enlarged before the year 2000, so update the `analysis` variable to only contain those rows

```{r}
analysis <- analysis |> filter(action == "Enlarged" & year < 2000)
```

#### Generate a summary by state and save it to a variable called `result` with the following statistics:

-   Total acres (i.e. sum) of the area enlarged
-   Time span in years (i.e. difference between first and last year the enlargement happened)
-   Number of times and area was enlarged (look up the `n()` function in dplyr for details)

```{r}
result <- analysis |> group_by(states) |> 
  summarise(
    area = sum(acres_affected), 
    timeSpan = max(year) - min(year), 
    nTimes = n()
  ) |>
  ungroup()
```

#### Add a column `large` to these results indicating whether the total area is larger than 100,000 acres

```{r}
result |> mutate(large = area >= 100000)
```
